---
title: "R Basics"
format:
  revealjs:
    scrollable: true
    navigation-mode: vertical
    controls-layout: bottom-right
    controls-tutorial: true
    incremental: false 
    chalkboard:
      src: chalkboard.json
      storage: chalkboard_pres
      theme: whiteboard
      chalk-width: 4
engine: knitr
knitr:
  opts_chunk: 
    echo: true
    eval: false
    comment: "#>" 


revealjs-plugins:
  - pointer
  - verticator
  
filters: 
  - reveal-header
  - code-fullscreen
  - webr

webr: 
  show-startup-message: true
---

```{r}
#| include: false
#| eval: true
#| label: setup

library(DT)
library(tidyverse)
```



# R

## R Programming

R is a st   atistical programming package that allows you to conduct different types of analysis.

[R](https://www.r-project.org/)

## RStudio

A piece of software that organizes how you conduct statistical analysis in R.

[RStudio](https://posit.co/downloads/)

## Posit Cloud

A web version of RStudio.

[Posit Cloud](https://posit.cloud/login)

## R Packages


::: panel-tabset

### Core Packages

-   Tidyverse

-   csucistats

```{r}
#| eval: false
#| echo: true
install.packages('csucistats', 
  repos = c('https://inqs909.r-universe.dev', 
  'https://cloud.r-project.org'))
```

### Course Packages


::: columns

::: column

- ISLR2

- leaps

- gam

- tree


:::

::: column

- gbm

- e1071

- glmnet

- torch

:::

:::

### Misc Packages

- RMarkdown

- Quarto

- extraInserts (GitHub Package)

:::

## Load R Package

```{r}
#| echo: true
#| eval: false

library(tidyverse)
```


- You must load the package every new R Session

# R Basics

## R as a calculator

R can evaluate different expressions in the console tab.

Try the following:

1.  $4(4+2)/34$
2.  $6^3$
3.  $3-1$
4.  $4+4/3+45(32*34-54)$

## R Calculator

```{webr-r}

```

## R Functions

R functions performs tasks to specific data values.

Evaluate the following values in R:

1.  $\sqrt{3}$
2.  $e^3$
3.  $\ln(53)$
4.  $\log(324)$
5.  $\sin(3)$
6.  $\sin(3\pi)$

## R Functions

```{webr-r}

```

## Types of Data

-   Numeric

-   Character

-   Logical

-   Missing

Evaluate the following code:

```{r}
#| eval: false

is.numeric(1)
is.numeric("1")
is.numeric(T)
is.numeric(NA)

```

## Types of Data

```{webr-r}

```


## Types of Objects

In R, an object contains a set of data. The most common types are `vectors` and `matrix`.

Run this code and print out the objects in the console:

```{r}
x <- 3:34
y <- matrix(1:20, nrow = 4)
```

## Types of objects

```{webr-r}

```


## Data Frames

Data frames can be thought of as R's version of a data set.

Play around with mtcars:

```{r}
#| eval: true
mtcars 
```

## Lists

List can be thought as an extended vector, but each element is a different R object.

Try playing with this R object:

```{r}
list_one <- list(mtcars, rep(0, 4),
                 diag(rep(1, 3)))
```

## Lists

```{webr-r}

```

