---
title: "Multivariable Linear Regression"
format:
  revealjs:
    scrollable: true
    navigation-mode: vertical
    controls-layout: bottom-right
    controls-tutorial: true
    incremental: false 
    chalkboard:
      src: chalkboard.json
      storage: chalkboard_pres
      theme: whiteboard
      chalk-width: 4
knitr:
  opts_chunk: 
    echo: false
    eval: false
    message: false
    warnings: false
    comment: "#>" 

editor: visual
---

```{r}
#| include: false
#| eval: true
#| label: setup

library(DT)
library(tidyverse)
library(GLMsData)
library(broom)
library(palmerpenguins)
library(magrittr)

x <- rnorm(50, 4)
z <- rnorm(50, -2)
y <- 3 + 2 * x + 4 * z + rnorm(50, sd = 2)
df <- tibble(x, y, z)



```

## Learning Objectives

-   Multivariable Linear Regression

-   Simulation Study 1

-   Simulation Study 2

-   Matrix Formulation

# Multivariable Linear Regression

## MLR

Multivariable linear regression models are used when more than one explanatory variable is used to explain the outcome of interest.

## Continuous Variable

To fit an additional continuous random variable to the model, we will only need to add it to the model:

$$
Y = \beta_0 +\beta_1 X_1 + \beta_2 X_2
$$

## Example

Looking at the `penguins` from `palmerpenguins`, fit a model between `body_mass_g` as an outcome variable and `flipper_length_mm` and `bill_length_mm` as predictor variables.

```{r}
#| echo: true
library(palmerpenguins)
penguins |>  lm(body_mass_g ~ flipper_length_mm + bill_length_mm, 
                data = _)
```

## Interpretation

```{r}
#| eval: true
library(palmerpenguins)
penguins |>  lm(body_mass_g ~ flipper_length_mm + bill_length_mm,
                data = _)
```

## Categorical Variable

A categorical variable can be included in a model, but a reference category must be specified.

## Fitting a model with categorical variables

To fit a model with categorical variables, we must utilize dummy (binary) variables that indicate which category is being referenced. We use $C-1$ dummy variables where $C$ indicates the number of categories. When coded correctly, each category will be represented by a combination of dummy variables.

## Example

If we have 4 categories, we will need 3 dummy variables:

|         | Cat 1 | Cat 2 | Cat 3 | Cat 4 |
|---------|-------|-------|-------|-------|
| Dummy 1 | 1     | 0     | 0     | 0     |
| Dummy 2 | 0     | 1     | 0     | 0     |
| Dummy 3 | 0     | 0     | 1     | 0     |

Which one is the reference category?

## Fitting a model with categorical variables

Looking at the `penguins` from `palmerpenguins`, fit a model between `body_mass_g` as an outcome variable and `flipper_length_mm` and `islands` as predictor variables.

```{r}
#| echo: true
library(palmerpenguins)
penguins |>  lm(body_mass_g ~ flipper_length_mm + island,
                data = _)
```

## Interpreting Categorical Variable Results

```{r}
#| eval: true
library(palmerpenguins)
penguins |> lm(body_mass_g ~ flipper_length_mm + island,
               data = _)
penguins |>  with(levels(island))
```

## Relevel a factor variable in R

```{r}
#| echo: true
penguins <- penguins |> 
  mutate(island_dream = relevel(island , ref = "Dream"))
penguins |> lm(body_mass_g ~ flipper_length_mm + island_dream, 
               data = _)
```

# Simulation Study 1

## Simulation Study 1

Simulate 1000 random variables from the following model:

$$
Y = 3 + 2X_1 + 4X_2 + \epsilon
$$

-   $X_1\sim N(2,1)$

-   $X_2\sim N(-4,1)$

-   $\epsilon\sim N(0, 2)$

```{r}
x1 <- rnorm(1000, 2)
x2 <- rnorm(1000, -4)
y <- 3 + 2 * x1 + 4 * x2 + rnorm(1000, sd = sqrt(2))
df <- tibble(x1, x2, y)
```

## Fit Model

Fit a model between $Y$ and $X_1$.

Repeat the process 1000 times. and answer the following questions:

-   On average does $\beta_1$ get estimated correctly? Why?

-   What is the average model variance?

## MLR Model

Instead of fitting a simple linear regression model. Fit a model that will include predictor $X_2$. This can be done by adding $X_2$ in R:

```{r}
#| echo: true

lm(y ~ x1 + x2)

```

Modify your simulation study and see what happens to $\beta_1$ and the model variance.

# Simulation Study 2

## Simulation Study 2

Simulate 1000 random variables from the following model:

$$
Y = 3 + 2log(X_1) + \epsilon
$$

-   $X_1\sim N(8,1)$

-   $\epsilon\sim N(0, 2)$

```{r}
x <- rnorm(1000, 8)
y <- 3 + 2 * log(x) + rnorm(1000, sd = sqrt(2))
df <- tibble(x, y)
lm(y~log(x))
```

## Fit Model

Fit a model between $Y$ and $X_1$.

Repeat the process 1000 times. and answer the following questions:

-   On average does $\beta_1$ get estimated correctly? Why?

-   What is the average model variance?

## SLR Model

Fit a simple linear regression model using $\log(X_1)$ instead.

```{r}
#| echo: true

lm(y ~ log(x1))

```

Modify your simulation study and see what happens to $\beta_1$ and the model variance.

# Matrix Formulation

## Matrix Formulation

$$
Y_i = \boldsymbol X_i^\mathrm T \boldsymbol \beta + \epsilon_i
$$

-   $Y_i$: Outcome Variable

-   $\boldsymbol X_i$: Predictors

-   $\boldsymbol \beta$: Coefficients

-   $\epsilon_i$: error term

## Matrix Formulation

Fit the following models using matrix formulas instead of the `lm` function.

$$
bmg = \beta_0+\beta_1 flipper\_length\_mm + \beta_2 bill\_length\_mm
$$

$$
bmg = \beta_0 +\beta_1 flipper\_length\_mm + \beta_2 dream + \beta_3 biscoe  
$$

$$
\boldsymbol\beta = (\boldsymbol X^\mathrm T \boldsymbol X)^{-1} \boldsymbol X^\mathrm T \boldsymbol Y
$$

Define $\boldsymbol X$ in each case.

# R Tips

## Create a Model Matrix X

$$
bmg = \\
\beta_0+ \beta_1bill\_length\_mm +  \beta_2 dream +\beta_3 biscoe  
$$

## Model Matrix

```{r}
#| echo: true
penguins |> model.matrix(~flipper_length_mm + bill_length_mm + islands, data = _)

```
