---
title: "Generalized Linear Models"
subtitle: "Model Inference"
format:
  revealjs:
    scrollable: true
    navigation-mode: vertical
    controls-layout: bottom-right
    controls-tutorial: true
    incremental: false 
    chalkboard:
      src: chalkboard.json
      storage: chalkboard_pres
      theme: whiteboard
      chalk-width: 4
knitr:
  opts_chunk: 
    echo: true
    eval: true
    message: false
    warnings: false
    comment: "#>" 
    
revealjs-plugins:
  - pointer
  - verticator
  
filters: 
  - reveal-header
  - code-fullscreen

editor: visual
---

## Learning Outcomes

-   Hypothesis Tests

-   Model Inference

# Hypothesis Tests

## Hypothesis Tests

Hypothesis tests are used to test whether claims are valid or not. This is conducted by collecting data, setting the **Null** and **Alternative** Hypothesis.

## Null Hypothesis

## Alternative Hypothesis

## Common Hypothesis Tests

-   t-tests

-   ANOVA

-   $\chi^2$-tests

## Testing $\beta_j$

::: columns
::: {.column width="50%"}
### $\phi$ known

$$
\frac{\hat\beta_j - \theta}{\mathrm{se}(\hat\beta_j)} \sim N(0,1)
$$
:::

::: {.column width="50%"}
### $\phi$ unknown

$$
\frac{\hat\beta_j-\theta}{\mathrm{se}(\hat\beta_j)} \sim t_{n-p^\prime}
$$
:::
:::

## Confidence Intervals

$$
PE \pm CV \times SE
$$

-   PE: Point Estimate

-   CV: Critical Value $P(X<CV) = 1-\alpha/2$

-   $\alpha$: significance level

-   SE: Standard Error

## Model Inference

$$
\hat\eta \pm CV\times SE
$$

$$
(LB = \hat\eta - CV\times SE, UB = \hat\eta + CV\times SE
$$

## Model inference for $\mu$

$$
\{g^{-1}(LB), g^{-1}(UB)\}
$$

## Applications in R

```{r}
library(survival)
library(tidyverse)
library(magrittr)
pois_glm <- bladder1 %$% glm(recur ~ treatment + number + size, 
                family = poisson)


```

## Individual Hypothesis Testing

```{r}
pois_glm %>% summary
```

## Model Inference

Expected Reoccurrence for those on Pyridoxine and 2 tumors, largest tumor 4 cm.

```{r}
newdata <- data.frame(treatment = "pyridoxine", number = 2, size = 4)
res <- pois_glm %>% predict(newdata, 
                     se.fit = T, 
                     type = "response") %T>% print
ci <- (res$fit + c(-1,1) * qnorm(0.975) * res$se.fit) %T>% print
exp(res$fit) # Expected occurrence
exp(ci) # 95% C
```

## Model Inference

For those who are on `thiotepa`, what is the difference in the average number of recuccrence for a patient who has 6 than a patient who has 0 tumors, keeping size constant.
