---
title: "Data Manipulation"
format:
  revealjs:
    scrollable: true
    navigation-mode: vertical
    controls-layout: bottom-right
    controls-tutorial: true
    incremental: false 
    chalkboard:
      src: chalkboard.json
      storage: chalkboard_pres
      theme: whiteboard
      chalk-width: 4
knitr:
  opts_chunk: 
    echo: true
    eval: false
    message: false
    comment: "#>" 

editor: visual
---

```{r}
#| include: false
#| eval: true
#| label: setup

library(DT)
library(tidyverse)
library(tidylog)
```

## Learning Objectives

-   `tidyr` Functions

-   Wide to Long Example

-   Plotting with `ggplot2`

# `Tidyr`

## `tidyr` Functions

A set of functions that will tidy up a data set such that:

-   Every Column is a variable

-   Every Row is an observation

-   Every Cell is a single value

## `pivot_longer()`

-   The `pivot_longer()` function grabs the variables that repeated in an observation places them in one variable

## `pivot_wider()`

-   The `pivot_wider()` function then converts long data to wide data.

## `separate()`

-   The `separate()` function will separate a variable to multiple variables:

# Example

## Wide to Long Data Example

We work on converting data from wide to long using the functions in the tidyr package. For many statistical analysis, long data is necessary.

## Load Data

Use the `read_csv()` to read `data_3_4.csv` into an object called `data1`;

```{r,message=F}
#| eval: true
data1 <- read_csv(file="http://www.inqs.info/files/hiss_3/data_3_4.csv")
```

## Wide Data

```{r}
#| echo: false
#| eval: true
datatable(data1, options = list(pageLength = 3))
```

## Long Data

```{r,include=FALSE}
#| eval: true
data1_long <- data1 %>% pivot_longer(`v1/mean`:`v4/median`,
                                     names_to = "measurement",
                                     values_to = "value") %>% 
          separate(measurement, c("time","stat"),sep="/") %>% 
          pivot_wider(names_from = stat,values_from = value)
```

```{r}
#| echo: false
#| eval: false
datatable(data1_long, options = list(pageLength = 8))
```

## `pivot_longer()`

-   The `pivot_longer()` function grabs the variables that repeated in an observation places them in one variable:

::: fragment
```{r}
#| code-line-numbers: "|2"
#| eval: true
#| message: true
df1 <- data1 %>% 
  pivot_longer(cols=`v1/mean`:`v4/median`,
               names_to = "measurement",
               values_to = "value")
```
:::

## `separate()`

-   The `separate()` function will separate a variable to multiple variables:

::: fragment
```{r}
#| code-line-numbers: "|3"
#| eval: true
#| message: true
df2 <- data1 %>% 
  pivot_longer(cols=`v1/mean`:`v4/median`,
               names_to = "measurement",
               values_to = "value") %>% 
  separate(col=measurement,into=c("time","stat"),sep="/")
```
:::

## `pivot_wider()`

-   The `pivot_wider()` function then converts long data to wide data.

::: fragment
```{r}
#| code-line-numbers: "|4"
#| eval: true
#| message: true
df3 <- data1 %>% 
  pivot_longer(`v1/mean`:`v4/median`,
               names_to = "measurement", 
               values_to = "value") %>% 
  separate(measurement,c("time","stat"),sep="/") %>% 
  pivot_wider(names_from = stat,
              values_from = value)
```
:::

# ggplot2

## ggplot2

`ggplot2` is an R package used to create plots. The main idea is to use a data frame and a set of aesthetics (variables in the data frame) to create a base plot. Then, `ggplot2` will layer geometries (plots) to the base plot to create a data visualization.

All new changes to the plot are layered on with the `+` symbol.

## Base Plot

```{r}
mtcars |> ggplot(aes(x = mpg)) 
```

## Histogram

```{r}
mtcars |> ggplot(aes(x = mpg)) +
  geom_histogram()
```

## Box Plot

```{r}
mtcars |> ggplot(aes(x = mpg)) +
  geom_boxplot()
```

## Density Plot

```{r}
mtcars |> ggplot(aes(x = mpg)) +
  geom_density()
```

## Box Plot By Category

```{r}
mtcars |> ggplot(aes(x = mpg, y = as.factor(cyl))) +
  geom_boxplot()
```

## Density Plot By Category

```{r}
mtcars |> ggplot(aes(x = mpg, color = as.factor(cyl))) +
  geom_density()
```

## Scatter Plot

```{r}
mtcars |> ggplot(aes(x = wt, y = mpg)) +
  geom_point()
```

## Scatter Plot by Group

```{r}
mtcars |> ggplot(aes(x = wt, y = mpg, color = as.factor(cyl))) +
  geom_point()
```

## Add Regression Line

```{r}
mtcars |> ggplot(aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = F)
```

## Smooth Line

```{r}
mtcars |> ggplot(aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(se = F)
```

## Regression Lines by Group

```{r}
mtcars |> ggplot(aes(x = wt, y = mpg,
                  color = as.factor(cyl))) +
  geom_point() +
  geom_smooth(method = "lm", se = F)
```

## Example

Using the `penguins` data set from `palmerpenguins` package. Create any plot and make it publication ready. Use the following resources to customize the plot: [R Graphics Cookbook](https://r-graphics.org/), [R Graph Gallery](https://r-graph-gallery.com/), [R Charts](https://r-charts.com/), and [ggplot2](https://ggplot2.tidyverse.org/)
