---
title: "Homework 2"
date: 2/7/24
editor: visual
format: 
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-tools: true
    code-line-numbers: true
    embed-resources: true
draft: false
---

**Due 2/17/24 \@ 11:59 PM**

Use an RMD M408 Template to create the assignment.

You must comment all your code to receive credit.

Submit the \*.html file to canvas.

## Problem 1

Write a function that takes a vector as an input an returns a labeled list with the mean, median, and variance of the vector.

```{r}
#| code-fold: true
#| code-summary: Answer

xsummary <- function(x){
  list(mean = mean(x),
       median = median(x),
       variance = var(x))
}

```

## Problem 2

Write a function for the following equation:

$$
f(x, y, z) =\left\{\begin{array}{cc}
x^2+\sqrt y & z  = 0 \\ 
x^2+2x+3+ \log(y) & z = 1
\end{array}\right.
$$

Include any error messages if necessary.

```{r}
#| code-fold: true
#| code-summary: Answer

fxyz <- function(x, y, z){
  if (!z %in% c(0,1)){stop("z must be 0 or 1")}
  if (y<=0){stop("y must be greater than 0")}
  if (z ==0){
    return(x^2+sqrt(y))
  } else {
    return(x^2 + 2*x + 3 + log(y))
  }
}

```

## Problem 3

Create a function that will give you the first `x` Fibonacci numbers. You must use a `for` loop.

```{r}
#| code-fold: true
#| code-summary: Answer

fib <- function(x){
  post <- c(0,1)
  for(i in 3:x){
    post <- c(post, fib[i-2] + fib[i-1])
  }
  return(post)
}

```

## Problem 4

Create a function that will randomly generate `n` numbers that are greater than `x` from a $Pois(\lambda)$.

```{r}
#| code-fold: true
#| code-summary: Answer

new_pois <- function(n, x, lambda){
  nn <- 0
  post <- c()
  while(nn < n){
    new <- rpois(1, lambda)
    if (new > x){
      post <- c(post, new)
      nn <- length(post)
    }
  }
  return(post)
}
```
