---
title: "Homework 3"
date: 2/14/24
editor: visual
draft: false
---

**Due 2/24/24 \@ 11:59 PM**

Use an RMD or QMD file to create the assignment.

You must comment all your code to receive credit.

Submit the \*.html file to canvas.

## Problem 1

Using the vectors below, use `sapply()` to determine if the numbers are even or odd.

```{r}
x <- 1:20
```

## Problem 2

Evaluate the following function:

$$
k(j(i(h(g(f(x))))))
$$

-   $f(x)=\sin(x)$

-   $g(x) = x^2 +5$

-   $h(x) = \exp{2x}$

-   $i(x) = \sin(\pi x)$

-   $j(x) = x^2 + 5$

-   $k(x) = \sqrt x$

You are not allowed to store values, but you may create and store functions.

## Problem 3

From the `penguins` data set from `palmerpenguins` package, create a new variable indicating if the penguin's flipper length (`flipper_length_mm`) is greater than the average flipper length. Afterwards, group by the new variable and find the mean `body_mass_g`.

## Problem 4

From the `penguins` data set from `palmerpenguins` package, find the mean and standard deviation of `flipper_length_mm` by `species`.

## Problem 5

Using the vector `random_dates` generated from the code below:

```{r}
#| eval: false
# Set the start and end dates
start_date <- as.POSIXct("2022-01-01 00:00:00")
end_date <- as.POSIXct("2022-12-31 23:59:59")

# Generate a vector of random dates and times between the start and end dates
random_dates <- as.POSIXct(runif(10, start_date, end_date), origin = "1970-01-01")

# Print the vector of random dates and times
random_dates

```

Use `sapply()` to determine what day of the week (in words) does the date represent. Hint: Look at the `wday` function from `lubridate`.
