---
title: "Homework 5"
date: 3/7/24
editor: visual
draft: false
---

**Due 3/17/24 \@ 11:59 PM**

Submit the \*.html file to canvas.

## Problem 1

The `faithful` data set in R contains information on `eruptions` time and `waiting` time. Describe the relationship between `waiting` (independent) and `eruptions` (dependent).

You must provide descriptive statistics, visuals, and model.

## Problem 2

When estimating the $\beta$ coefficients, we are minimizing the sum of error squares:

$$
\sum^n_{i=1}(y_i-(\beta_0+\beta_1x_i))^2
$$

We square the errors to make sure we do not lose information when summing up all the values. However, what if we try to use an absolute value:

$$
\sum^n_{i=1}|y_i-(\beta_0+\beta_1x_i)|
$$

Simulate a data set, estimate the coefficients, and compare the results. Are they the same or different?

## Problem 3

Conduct a simulation assessing which case from problem 3 is better.

To assess which case is better compare the mean and standard deviations of the simulated-estimated coefficients.

## Problem 4

Run the code below to obtain the data:

```{r}
#| echo: true
#| eval: false

library(tidyverse)
anime <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-04-23/tidy_anime.csv") |> filter(!type == "Unknown") |> as.data.frame()

```

The `anime` data set from MyAnimeList contains information on rankings and popularity scores of different anime episodes.

Fit a linear model showing the relationship between `score` (outcome; higher is better) and the predictors `type`, `popularity`, and `rank`. On average, what is the score for an anime that is a "Special", popularity score of 520, and rank of 3582?

## Problem 5

Run this code below to obtain the data (you may need to install `tidytuesdayR`):

```{r}
#| echo: true
#| eval: false

library(tidyverse)
tuesdata <- tidytuesdayR::tt_load(2021, week = 52)
starbucks <- tuesdata$starbucks |> drop_na() |> filter(size %in% c("tall", "grande", "venti"))

```

Fit a linear model between the outcome `caffeine_mg` (outcome) and predictors `size`, `calories`, and `sugar_g` from the `starbucks` data set using the matrix formulation approach (do matrix algebra) by defining the $X$ matrix. Use `tall` as the reference value for the `size` variable.
