---
title: "Class 8"
author: "Isaac Quintanilla Salinas"
date: "`r format(Sys.time(),'%m-%d-%Y')`"
output: 
  html_notebook:
    toc: true
    toc_float: true
    code_folding: hide
---

## Anonymous Example 1

Use an anonymous function to square all the values in the following vector:

```{r}
# Use an anonymous function to calculate the square of each element in a vector
numbers <- 1:40
sapply(numbers, \(y) y^2)
numbers^2
```

## 2Anonymous Example 2

Use an anonymous function to convert the vector from Fahrenheit to Celsius:

```{r}
# Create a vector of temperatures in Fahrenheit
temperatures_f <- c(32, 68, 104, 50)
sapply(temperatures_f, function(f) 5*(f-32)/9)
```

$$
C = \frac{5(F-32)}{9}
$$

## Pipe Examples 1

Using the vector below, find the standard deviation using a pipe:

```{r}
library(magrittr)
x <- rgamma(100, 1)
sd(x)
x %>% sd
```

## Pipe Examples 2

Chain pipe the previous results into the $sin(x)$.

```{r}
x %>% sd %>% sin
```

## Pipe Examples 3

Chain pipe the previous results into $e^x$.

```{r}
x %>% sd %>% sin %>% exp
```

## Pipe Examples 4

Chain pipe the previous results into $x^2+5x+4$

```{r}
x %>% sd %>% sin %>% exp %>% (\(x) x^2 +5*x+4)
```

```{r}
library(palmerpenguins)
penguins
```
