---
title: "Class 7"
author: "Isaac Quintanilla Salinas"
date: "`r format(Sys.time(),'%m-%d-%Y')`"
output: 
  html_notebook:
    toc: true
    toc_float: true
    code_folding: hide
---

## `apply()`

### Example

Find the standard deviation of all the columns of the following matrix:

```{r}
x <- matrix(rnorm(1000), nrow = 10)
apply(x, 2, sd)
```

### Example

Find the $25th$, $50th$, and $75th$ quartiles for each row of the following matrix:

```{r}
x <- matrix(rnorm(1000), nrow = 20)
apply(x, 1, quantile, probs = c(0.25, 0.50, 0.75))
```

## `lapply()`

## Example

Create a function that returns a labeled list for with the following values: mean, standard deviation, median, max, and min.

```{r}
sum_x <- function(x){
  x_list <- list(mean = mean(x),
                 sd = sd(x),
                 median = median(x),
                 min = min(x),
                 max = max(x))
  return(x_list)
}
sum_x(rnorm(100))
```

## Example

```{r}
lists <- lapply(1:10, function(i) {
  means <- rpois(1, 3)
  rnorm(5, means)
})

lapply(lists, sum_x)
```

## `sapply()`

## Example

Using the vector below, compute the length of each string using `sapply` and `str_length()` from `tidyverse`

```{r}

fruits <- c("apple", "banana", "cherry", "date", "elderberry", "fig", "grape", "honeydew", "kiwi", "lemon")
library(tidyverse)
sapply(fruits, str_length)
```

## Example

Using the list generated below, compute the mean of each element of the list using `sapply`.

```{r}
# Generate a list of 10 lists, each containing 5 random numbers
lists <- lapply(1:10, function(i) {
  means <- rpois(1, 3)
  rnorm(5, means)
})
sapply(lists, mean)
```

## Example

Using the vector below, use the `sapply()` to find $\log(x)$ for each value and return a matrix:

```{r}
numbers <- 4:400
sapply(numbers, log, base = 10, simplify = "matrix")
```

## `mapply()`

## Example

Let `x` and `y` be two vectors, shown below, represent the x and y coordinates of a point. Using `mapply()`, compute the the distance between the points and the origin.

```{r}
x <- c(2, 3, 4, 5)
y <- c(4, 6, 8, 10)

pyth <- \(x,y){
  sqrt(x^2+y^2)
}

mapply(pyth, x=x, y=y)
```

## `tapply()`

## Example

Using the `penguins` data set from the `palmerpenguins` package, compute the average `bill_length_mm` for each `island`.

```{r}

#install.packages("palmerpenguins")
library(palmerpenguins)
names(penguins)
tapply(penguins$bill_length_mm, 
       penguins$island, mean, na.rm = T)
```

## Example

The vectors below provide the heights of different trees in the sample. Compute the median for each type of tree.

```{r}
heights <- c(70, 72, 68, 65, 80, 75, 60, 68, 90, 72)
species <- c("maple", "oak", "pine", "maple", "oak", "pine", "maple", "oak", "pine", "maple")
tapply(heights, species, median)
```
