---
title: "Class 9"
author: "Isaac Quintanilla Salinas"
date: "`r format(Sys.time(),'%m-%d-%Y')`"
output: 
  html_notebook:
    toc: true
    toc_float: true
    code_folding: hide
---

## Reading Data Example

```{r}
library(tidyverse)
data1 <- read_csv("/home/inqs/Repos/M408_S23/files/data/data_3_1.csv")
data2 <- read_csv("/home/inqs/Repos/M408_S23/files/data/data_3_2.csv")
```

## Example

Load data `data_3_1.csv` and `data_3_2.csv`.

```{r}
data1
data2
```

## Example

Load the following data: <https://m408.inqs.info/files/data/data_3_3.csv>

```{r}

```

## Example

Merge data sets `data_3_1.csv` and `data_3_2.csv` using the `full_join()`

```{r}

```

## Example

Using the `penguins` dataset from `palmerpenguins`, create a new variable that is the ln of `flipper_length_mm`.

```{r}

```

## Example

Using the `penguins` dataset from `palmerpenguins`, only select the variables that are continuous data points.

```{r}

```

## Example

Using the `penguins` dataset from `palmerpenguins`, filter the data set to look at penguins that are a Gentoo species.

```{r}

```

## Example

Using the `penguins` dataset from `palmerpenguins`, create a new variable that dichotomizes a penguin if their bill is longer than the average `bill_length_mm`.

```{r}

```

## Example

Using the `penguins` dataset from `palmerpenguins`, group by species and find the average ln `flipper_length_mm`

```{r}

```
